---
title: "Analysis of Variance"
author: "Gaëlle Cordier"
date: "24/11/2014"
output: html_document
---

---

```{r figures_path,echo=FALSE,message=FALSE}
library(knitr)
opts_chunk$set(fig.path = "figures/") # Set figures path
```

```{r load_libraries,echo=FALSE,message=FALSE}
library(reshape2)
library(doBy)
library(ggplot2)
library(gridExtra)
library(gtable)
```

ANOVA compares (two or more) means by comparing variances; one-Way ANOVA compares (two or more) means based on one factor by comparing variances.

Data example: crop yields per unit area measured from 10 randomly selected fields on each of 3 soil types (sand, clay, and loam) 

```{r data_yields}
yields<-data.frame(sand=c(6,10,8,6,14,17,9,11,7,11),
                   clay=c(17,15,3,11,14,12,12,8,10,13),
                   loam=c(13,16,9,12,15,16,17,13,18,14))
yields
```

We may want to know whether soil types (categorical explanatory variable or factor) significantly affects crop yield (numerical respone variable):

$H_0: \mu_s = \mu_c = \mu_l$

$H_1: \text{at least one mean is different}$

# Test yield normality and yield~soils variance homogeneity (heterocedasticity)!

If we take a look at the sample means:

```{r reshape_data}
yields_l<-melt(data = yields,
               value.name = "yield",
               measure.vars = c("sand","clay","loam"),
               variable.name = c("soil"))

head(yields_l); tail(yields_l)
```

```{r soil_means}
means<-summaryBy(formula = yield~soil,data = yields_l,FUN = mean)
means
# or:
# aggregate(yield~soil, yields_l, mean)
```

and at the distribution of the yield values within the 3 soils:

```{r plot_soils}
ggplot(data = yields_l,aes(x = soil,y = yield, color = soil))+
    geom_boxplot()+
    stat_summary(fun.y=mean, geom="point", shape=4)+
    labs(x="")+
    scale_color_discrete(guide=F)
```

we can see that yield may turn out to be significantly different between sand and loam soils (their boxes don't overlap), but it is not as clear whether clay yield will be significantly greater/lower than sand/loam yield.

How does ANOVA allows us to make inferences about differences between means by looking at differences between variances?

### Deviation (variability) and variance:

---

The distance from any point in a collection of data, to the mean of the data (sample mean), is the deviation. This can be written as:

$y_i-\bar{y}$

where $y_i$ is the *ith* data point, and $\bar{y}$ is the estimate of the mean. If all such deviations are squared (so all differences are positives), and then summed, as in:

$\sum_{i=1}^n\left(y_i-\bar{y}\,\right)^2$

this gives the "sum of squares" for these data.

The deviation is unscaled (the sum of squares will grow with the size of the data collection), so to compare samples of different sizes we need to scale it by dividing it by the degrees of freedom (the number of parameters of the system that may vary independently, or to simplify, the sample size minus 1). In fact, the sample variance of a discrete random variable is defined as:

$S^2=\frac{\sum_{i=1}^{n}(y_i-\bar{y})^2}{n-1}$

So, the deviation is an unscaled measure of dispersion (or variability), that when scaled for the number of degrees of freedom estimates the variance.

### Partition of Sum of Squares:

---

The analysis of variance involves calculating the total variation or variability in the response variable (yield in this case) and partitioning it into two components: the intra-groupal or unexplained variability (variability in each group due to unknown factors) and the inter-groupal or explained variability (variability due to the explanatory variable, soil in this case). This is called the partition of sum of squares, and it allows us to quantify the relative importance of each one of said sources of variability: in our case, if the factor soil has an effect over crop yield, we would expect the total variability to be explained in greater measure by the explained variability than by the unexplained variability:

$\text{TOTAL VARIABILITY = EXPLAINED VARIABILITY + UNEXPLAINED VARIABILITY}$

### Calculating the Sums of Squares

---

#### **SSE**

We can define the variability in the response variable within each group as:

$\text{SSE}=\sum_{i=1}^{k}\sum_{i=j}^{n}(y_{ij}-\bar{y})^2$

that is, the sum of squares of the differences between the observation *j* (*n=10* replicates) within each group and the mean of said group *i* (*k=3* factor levels). This would be the unexplained variation or residual variability (error sum of squares) since it is not explained by the differences between groups.

```{r SSE}
yields_l2<-data.frame(yields_l,mean=rep(means$yield.mean,each = 10))

yields_l2

SSE<-with(data = yields_l2,expr = sum((yield-mean)^2))

SSE
```

$\\$

#### **SSA**

In a similar way, we can define the variability in the response variable between groups as:

$\text{SSA}=\sum_{i=1}^{k}\sum_{i=j}^{n}(\bar{y_{i}}-\bar{\bar{y}})^2=n*\sum_{i=1}^{k}(\bar{y_{i}}-\bar{\bar{y}})^2$

that is, the sum of squares of the differences between the individual treatment means *n x i* and the overall mean (the mean of all observations, or the mean of the group means). This would be the explained variability (the treatment sum of squares).

```{r SSA}
means

SSA<-10*sum((means$yield.mean-mean(means$yield.mean))^2)

SSA

# or
with(data = yields_l2,expr = sum((mean-mean(yields_l2$yield))^2))
```

$\\$

#### **SST**

The total variability (the total sum of squares) would be then:

$\text{SST}=\sum_{i=1}^{k}\sum_{i=j}^{n}(y_{ij}-\bar{\bar{y}})^2$

that is, the sum of squares of the differences between the observation *ij* and the overall mean.

```{r SST}
SST<-with(data = yields_l2,expr = sum((yield-mean(yield))^2))

SST
```

$\\$

**Plotting the variability**

```{r SSE_plots}
plotdata<-data.frame(yields_l2,
                     ov.mean=rep(mean(means$yield.mean)),
                     x=seq(from = 0,to = 30,length.out = nrow(yields_l2)))

p1<-ggplot(data = plotdata,aes(x=x,y=yield,shape=soil,color=soil))+
    geom_point(size=2)+
    labs(x="",y="",title="Observations")
    # if we want to make legend horizontal:
#     guides(color=guide_legend(direction="horizontal",title.position="top",title.hjust=0.5),
#            shape=guide_legend(direction="horizontal",title.position="top",title.hjust=0.5))

p2<-ggplot(data = plotdata,aes(x=x,y=yield,shape=soil,color=soil))+
    geom_point(size=2)+
    geom_line(aes(y=mean))+
    geom_linerange(aes(ymin=yield,ymax=mean))+
    labs(x="",y="",title="SSE")

p3<-ggplot(data = plotdata,aes(x=x,y=yield,shape=soil,color=soil))+
    geom_point(size=2)+
    geom_line(aes(y=mean),size=1)+
    geom_abline(intercept=11.9,slope=0)+
    geom_linerange(aes(ymin=mean,ymax=ov.mean))+
    labs(x="",y="",title="SSA")

p4<-ggplot(data = plotdata,aes(x=x,y=yield,shape=soil,color=soil))+
    geom_point(size=2)+
    geom_abline(intercept=11.9,slope=0)+
    geom_linerange(aes(ymin=yield,ymax=ov.mean))+
    labs(x="",y="",title="SST")

legend<-gtable_filter(ggplot_gtable(ggplot_build(p1)), "guide-box") # extract legend from p1

grid.arrange(arrangeGrob(p1 + theme(legend.position="none"), 
                         p2 + theme(legend.position="none"),
                         p3 + theme(legend.position="none"),
                         p4 + theme(legend.position="none"), 
                         nrow = 2,
                         main = textGrob("Sums of Squares", vjust = 1, gp = gpar(fontsize=15)),
                         left = textGrob("Yield", rot = 90, vjust = 1)),
             legend,ncol=2,widths=c(4/5,1/5))
```

*(extraction of legend and grid.arrange with global Y-axis and common legend as seen [here](http://stackoverflow.com/questions/11076567/plot-a-legend-and-well-spaced-universal-y-axis-and-main-titles-in-grid-arrange))*

$\\$

**In conclusion**

The total sum of squares is the sum of the treatment sum of squares and the error sum of squares:

$SST = SSA + SSE$

```{r check_SST}
all.equal(SST,SSA+SSE)
```

So the difference between SST and SSE is the treatment sum of squares, SSA:

$SSA = SST - SSE$

This is the amount of the variation in yield that is explained by the differences between the treatment means.

### Drawing the ANOVA table: the F-test

---

**Source of variation**

- Explained variability: type of soil
- Unexplained variability: error
- Total variability

```{r var_source}
Source<-c("Soil type","Error","Total")
```

$\\$

**Sum of squares**

- SSA
- SSE
- SST

```{r SS}
SS<-c(SSA,SSE,SST)
```

$\\$

**Degrees of freedom**

The degrees of freedom are the number of parameters that may vary independently (the number of observations minus 1).

- Treatment degrees of freedom: we are comparing 1 mean per soil type with the overall mean (3 parameters), so we have:   
$3-1=2$ degrees of freedom

- Error degrees of freedom: we are comparing 10 observations per soil type with the soil type mean (10 parameters by soil type), so we have:   
$(10-1)*3=9*3=27$ degrees of freedom

- Total degrees of freedom: we are comparing 30 observations with the overall mean (30 parameters), so we have:   
$30-1=29$ degrees of freedom

```{r df}
df<-c(2,27,29)
```

$\\$

**Mean square**

The mean square is obtained by divinding the sum of squares by the degrees of freedom, and is a measure of the treatment variance and the error variance:

```{r MS}
MS<-SS/df
```

The treatment variance is the mean square between groups ($MS_B$):

```{r MS_B}
MS_B<-MS[1]
MS_B
```

The error variance is the mean square error ($MSE$) or mean square within groups ($MS_W$), and since there is equal replication in each soil type, it is equal to the mean of the variances of the soil types:

```{r MS_W}
MS_W<-MS[2]
MS_W
vars<-aggregate(yield~soil, yields_l, var)
vars
mean(vars$yield)
```

The total variance is the total mean square, and it is equal to the variance of all the observations:

```{r}
MS_T<-MS[3]
MS_T
sum((yields_l2$yield-11.9)^2)/29
# or
var(yields_l2$yield)
```

$\\$

**F ratio**

$F=\frac{MS_B}{MS_W}$

The *F* ratio tests the null hypothesis that the treatment means are all the same:

$H_0: \mu_s = \mu_c = \mu_l$

$H_1: \text{at least one mean is significantly different from the others}$

If the null hypothesis isn't true, we would expect the $MS_B$ to be greater than the $MS_W$, so we expect the F-ratio to be > 1. On the contrary, if the null hypothesis is true, we expect the F-ratio to have a value close to 1.

```{r F_ratio}
FR<-MS_B/MS_W
FR
```

Is this value (> 1) significant? To answer this we must compare the test statistic F=4.24 with the critical value of F, that is, the value in the distribution from which we would be rejecting the null hypothesis. This critical value is then the quantile of the probability distribution for a given probability (0.95 if $\alpha=0.05$), given the degrees of freedom of the numerator (df=2) and the denominator (df=27):

$\text{formula quantile here}$

```{r qf}
qf(p = 0.95,df1 = 2,df2 = 27)
```

As the F-test > critical value, we would reject the null hypothesis. But in order to be able to work independently from the confidence interval, we must use the cumulative distribution, that is, we are looking for the probability of being equal or greater than our quantile; that would be the inverse of the cumulative distribution: 

$P(x \geq Ft) = 1 - P(x \leq Ft)$

```{r pf}
p_value<-1-pf(q = FR,df1 = 2,df2 = 27)
p_value
```

This is the p-value, that is, the probability for a value of being equal or greater than 4.24 when the null hypothesys is true; in this case, the probability of a value being equal or greater than 4.24 by chance would be 0.025 (25 times in 1000). If we are rejecting at a confidence level of 5% ($\alpha=0.05$), we would reject the null hypothesis since $p-value < \alpha$.

$\\$

**ANOVA table**

```{r AOV_table}
AOV.table<-data.frame(Source,SS,df,MS,FR,p_value)
AOV.table
```

This can be done fitting our analysis of variance model with `aov`, and then calling the `summary`:

```{r summary_aov}
yields.aov<-with(yields_l,aov(formula = yield~soil))
summary(yields.aov)
```

$\\$

**In conclusion**

We reject the $H_0: \mu_s = \mu_c = \mu_l \rightarrow$ there are significant differences between soil types (at least one mean is different).

### 

The same ANOVA table can be drawn with the `anova` function after fitting the model with `lm`:

```{r anova_lm}
yields.lm<-with(yields_l,lm(formula = yield~soil))
anova(yields.lm)
```

If we do the summary of the fit, we have:

```{r summary_lm}
sum.lm<-summary(yields.lm)
sum.lm$coefficients
```

The main difference between the `aov` and the `lm` approach is the way `summary` handles the fit: 

* `summary(fit.aov)` expresses the fit in terms of analysis of variance
* `summary(fit.lm)` expresses the fit in terms of linear models (the intercept, the slope, and their standard errors)

In the context of aov, the *Intercept* is a mean value; in this case, it corresponds to the mean of the first factor level (*soilsand*). The slopes are differences between means; each one corresponds to the difference between the means of each other factor (*soilclay* and *soilloam*) and the mean of the first one on the Intercept.

$\\$

**Estimate**

The estimate for Intercept is the mean of *soilsand*, the estimate for soilclay is the difference between the means of *soilclay* and *soilsand*, and the estimate for soilloam is the difference between the means of *soiloam* and *soilsand*:

```{r estimates}
sum.lm$coefficients[,1]

# means
means
# sand mean
means$yield.mean[1]
# clay mean - sand mean
means$yield.mean[2]-means$yield.mean[1]
# loam mean - sand mean
means$yield.mean[3]-means$yield.mean[1]
```

$\\$

**Standard error**

The standard deviation of a population, $\sigma$ (the square root of the variance population, $\sigma^2$), quantifies the deviation of individuals with respect to the true (the population) mean, $\mu$:

$\sigma = \sqrt{\sigma^2} = \sqrt{\frac{\sum_{i=1}^n (y_i - \mu)^2}{n}}, {\rm \ \ where\ \ } \mu = \frac{\sum_{i=1}^n y_i}{n}$

Its estimate $S$, the standard deviation of a sample, quantifies the deviation of individuals with respect to the sample mean, $\bar{y}$:

$S = \sqrt{S^2} = \sqrt{\frac{\sum_{i=1}^{n}(y_i-\bar{y})^2}{n-1}}$

The standard error of the mean, $SE_\text{mean}$, quantifies the deviation of a sample mean with respect to its true mean:

$SE_\text{mean} = \sqrt{\frac{S^2}{n}}$

The standard error on the Intercept is the standard error of the sand soil mean, where the variance is the error variance (the $MSE$ or $MS_W$), and *n* is the number of observations or replications in that sample (group):

$SE_\text{mean} = \sqrt{\frac{MS_W}{n}}$

```{r SE_mean}
sqrt(MS_W/10)
sum.lm$coefficients
```

It is a measure of how accurate our estimate of the mean is likely to be, given the existing variability within groups.